HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-04 submission.Rmd
Add changes from Rouven...

@@

##### 3.2 Enhanced post-motion model
We re-train the post-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as features. An outlet of this problem is to employ suitable summary statistics (`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus; Robert: Im moment nehme ich alle 4 (sum, average, nr changes, average ent)... Einfach alle bitte kurz beschreiben.","orange")`). In particular, we calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$. From this, we count the number of state switches and sum these numbers across defenders as a first, simple, summary statistics and further calculate across defenders $j = 1, \ldots, 5$ the mean entropy in offensive play $n = 1,\ldots, N$, i.e.:

@@

@@

While we were clearly able to show that motion enables the defensive coverage scheme, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. `r colorize("Noch mehr Fokus auf Modelle und der Möglichkeit flexibler Modelle, wie die von AMAZON zu fitten, wenn mehr Daten dabei sind.","orange")`). Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by our extention. `r colorize("Vergleiche unser Modell mit Results von Amazon: Unser Modell wesentlich weniger aufwändig, performt aber gut (Accuracy > 88 %. Accuracy von deren Modellen zwischen 88 und 96% obwohl sie 3 ganze Seasons an Daten haben (verglichen zu unseren 9 Wochen, wo wir 2/3 der plays entfernen müssen!)","orange")`).

>>>>

In the last step, we aim to use results from a hidden Markov model as further features in the previously described models. For this, we describe the movements of the five defensive players during the phase of pre-snap motion by an HMM (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a particular Gaussian distribution with mean according to the $y$-coordinate of the offensive player that is currently guarded and an estimated standard deviation. HMMs are particularly useful in this setting as there is no direct knowledge of the guarding allocation. This is only partially observed based on the reaction of the defender to the offenders' movement. Naturally, the model thus treats the information on who is currently guarded as a latent state variable, and hence each observation as a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
Ultimately, the goal is to use the fitted model for so-called *state-decoding*, i.e. inferring information on the guarding allocation based on the observations made. For this task, HMMs not only leverage information on the $y$-coordinates of the defender and all offenders potentially guarded at each time point, but also incorporate probabilistic information based on the previous and subsequent observations. For practical details on the model fitting and state decoding, also see the Appendix.

@@

##### 3.2 Enhanced post-motion model
We re-train the post-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as features. An outlet of this problem is to employ suitable summary statistics (`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus; Robert: Im moment nehme ich alle 4 (sum, average, nr changes, average ent)... Einfach alle bitte kurz beschreiben.","orange")`). To overcome this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$ and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $s$ and $t$, $s>t$. From this, we can calculate some simple, summary statistics: 1) the sum of state switches, 2) the average number of state switches and 3) the number of defenders during a play that switch offensive players during a play. Besides, we use the aforementioned state decodings to calculate a more elaborated statistic, in particular the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$, i.e.:

@@

When performing state decoding, it becomes evident that the HMM effectively captures Marco Wilson's guarding assignment. The unique strength of HMMs is particularly apparent when Marco Wilson reaches the same height as the running back, Jerick McKinnon (KC #1). At this point, clustering algorithms that disregard the temporal component would briefly suggest a change in the guarding assignment, introducing noise into summary statistics such as the total number of switches predicted by the model. In contrast, the HMM consistently and accurately maintains the correct coverage assignment throughout the motion.

<center>
  <img src="coverage.gif" width="500" height="500" alt="Coverage Prediction">
</center>

@@

While we were clearly able to show that motion facilitates the defensive coverage scheme detection, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. `r colorize("Noch mehr Fokus auf Modelle und der Möglichkeit flexibler Modelle, wie die von AMAZON zu fitten, wenn mehr Daten dabei sind.","orange")`). Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by our extension. `r colorize("Vergleiche unser Modell mit Results von Amazon: Unser Modell wesentlich weniger aufwändig, performt aber gut (Accuracy > 88 %. Accuracy von deren Modellen zwischen 88 und 96% obwohl sie 3 ganze Seasons an Daten haben (verglichen zu unseren 9 Wochen, wo wir 2/3 der plays entfernen müssen!)","orange")`).
